Surface correlation functions calculations: improving methodology for discrete images with novel exact continuous approach

### 2023-03-24 paper.tex
Add precise algorithm for F_{sss}

\subsection{Surface-surface function for two-dimensional sets with smooth boundary}
\label{sec:fss-2d}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/Fss.png}
  \caption[]{Illustration of how $F_{ss}(\bm{r})$ is calculated. The interface
    between solid and void phases is a solid black curve. This interface
    translated by the vector $\bm{r}$ is show as a red solid curve. Support of
    $f_\Gamma(\bm{x}; \Delta)$ is a dashed ring. Shaded regions contribute to
    the surface-surface function.}
  \label{fig:Fss-explained}
\end{figure}
Let $S$ be some set in $\mathbb{R}^2$ which has a smooth boundary $\Gamma$ in
the sense that the boundary has a tangent line almost everywhere. Now we
introduce a function $f_\Gamma(\bm{x}; \Delta)$:
\begin{equation*}
  f_\Gamma(\bm{x}; \Delta) = \left\{
  \begin{array}{ll}
    1/\Delta & \quad \rho(\bm{x}, \Gamma) < \Delta \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation*}
Here $\rho(\bm{x}, \Gamma)$ means a closest distance from point $\bm{x}$ to the
boundary $\Gamma$.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/fss-zoomed.png}
  \caption[]{Zoomed image of intersection between the interface $\Gamma$ and its
    translated version $\Gamma'$. $\phi$ is an angle between tangent lines to
    the interfaces. Surface of dashed area is
    $\frac{\Delta^2}{\sin \phi}$. Vectors $\bm{a}$ and $\bm{b}$ are unit vectors
    parallel to normals to $\Gamma$ and $\Gamma'$ respectively in the point of
    their intersection.}
  \label{fig:fss-zoomed}
\end{figure}
Now define a functional sequence $F_{ss}(\bm{x}; \Delta)$:
\begin{align*}
  F_{ss}(\bm{r}; \Delta) &= \int\int f_\Gamma(\bm{x}; \Delta) f_\Gamma(\bm{x}
  + \bm{r}; \Delta) dx dy \\
  &= \sum_{k=1}^N S_k/\Delta^2
\end{align*}
Here $N$ is a number of intersections of $\Gamma$ with itself translated by a
vector $\bm{r}$ and $S_k$ is a surface of each such intersection (see
\cref{fig:Fss-explained}).
When $\Delta$ tends to zero, the boundary at the point of intersection can
be replaced with a tangent line at the point of intersection
(\cref{fig:fss-zoomed}). We get an expression for a surface of
intersection: $S_k = \frac{\Delta^2}{\sin \phi_k}$ where $\phi_k$ is an
acute angle between tangent lines. Hence, we get the following equation:
\begin{equation}
  F_{ss}(\bm{r}) = \lim_{\Delta \to 0} F_{ss}(\bm{r}; \Delta) =
  \sum_{k=1}^N \frac{1}{\sin \phi_k} \label{eq:fss-2d-sin}
\end{equation}
It's easy to show that
$\lim_{\Delta \to 0} f_\Gamma(\bm{x}; \Delta) = |\nabla \chi_A(\bm{x})|$
and therefore \cref{eq:fss-2d-sin} and \cref{eq:fss} define the same function.

A less illustrative but more practical way to calculate an expression
$\frac{S_k}{\Delta}$ is the following. Take two vectors $\bm{a}$ and $\bm{b}$
parallel to normals to $\Gamma$ and $\Gamma'$ in $k$-th point of their
intersection such as $|a| = |b| = 1$. If $F_{ss}$ is defined at the point of
consideration then $\bm{a} \ne k\bm{b}$. The pair $(a, b)$ may serve as a basis
with the following transformation from the basis $(0,1), (1,0)$:
\begin{align*}
  x' &= f(x, y) = a_x x + a_y y \\
  y' &= g(x, y) = b_x x + b_y y
\end{align*}
where $\bm{a} = (a_x, a_y)$, $\bm{b} = (b_x, b_y)$. Jacobian determinant for
this transformation is
\begin{equation*}
  J = \left|
  \begin{array}{cc}
    \frac{\partial f}{\partial x} & \frac{\partial f}{\partial y} \\
    \frac{\partial g}{\partial x} & \frac{\partial g}{\partial y} \\
  \end{array}
  \right| =
  \left|
  \begin{array}{cc}
    a_x & a_y \\
    b_x & b_y
  \end{array}
  \right| = a_x b_y - a_y b_x
\end{equation*}

@@

Now we get to practical algorithm for computation of $F_{ss}$ for sets with a
smooth boundary. Suppose a set $S$ is defined as
$S = \left\{ (x, y) \ | \quad f(x, y) \le T \right\}$ where $f(x, y)$ is some
function differentiable at points where $f(x, y) = T$ and $T$ is some
threshold. Let's calculate surface-surface function for this set at a point
$(a, b)$. We find all intersections between a boundary $\Gamma$ and its
translated version by solving a system of equations
\begin{equation*}
  \left\{
  \begin{array}{l}
    f(x, y) = T \\
    f(x-a, y-b) = T    
  \end{array}
  \right.
\end{equation*}
which can be replaced by one equation
\begin{equation}
  (f(x, y) - T)^2 + (f(x-a, y-b) - T)^2 = 0 \label{eq:inter}
\end{equation}
We denote a set of solutions of this equation as $X$. For every $\bm{x}_i \in X$
we find normals to the boundary and the boundary translated by a vector
$(a, b)$. These normals can be found using automatic differentiation and dual
numbers described in \cref{sec:dual}. Then we normalize these normals (i.e. make
their length equal to $1$) and use \cref{eq:fss-2d} to calculate $F_{ss}$.

Summarizing, the algorithm of calculating $F_{ss}(a, b)$ is the following:
\begin{enumerate}
\item Find $X$ which is a set of soulutions of \cref{eq:inter}.
\item For all points in $X$ find normals to the boundaries using
  \cref{eq:autonormals}.
\item Calculate $F_{ss}(a, b)$ using \cref{eq:fss-2d}.
\end{enumerate}

The set $X$ can be found by firstly evaluating $f(x, y)$ in some regular grid
searching for all points $(x_i, y_i)$ for which $|f(x_i, y_i) - T| < \epsilon$
and $|f(x_i - a, y_i - b) - T| < \epsilon$ for some positive $\epsilon$
and then using these points as starting points for some algorithm based on
gradient descent like ADAM to solve \cref{eq:inter}.

>>>>

\subsection{Surface-surface function for two-dimensional sets with smooth boundary}
\label{sec:fss-2d}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/Fss.png}
  \caption[]{Illustration of how $F_{ss}(\bm{r})$ is calculated. The interface
    between solid and void phases is a solid black curve. This interface
    translated by the vector $\bm{r}$ is show as a red solid curve. Support of
    $f_\Gamma(\bm{x}; \Delta)$ is a dashed ring. Shaded regions contribute to
    the surface-surface function.}
  \label{fig:Fss-explained}
\end{figure}
Let $S$ be some set in $\mathbb{R}^2$ which has a smooth boundary $\Gamma$ in
the sense that the boundary has a tangent line almost everywhere. Now we
introduce a function $f_\Gamma(\bm{x}; \Delta)$:
\begin{equation}
  f_\Gamma(\bm{x}; \Delta) = \left\{
  \begin{array}{ll}
    1/\Delta & \quad \rho(\bm{x}, \Gamma) < \Delta \\
    0 & \quad \text{otherwise}
  \end{array}
  \right. \label{eq:delta-sequence}
\end{equation}
Here $\rho(\bm{x}, \Gamma)$ means a closest distance from point $\bm{x}$ to the
boundary $\Gamma$.

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/fss-zoomed.png}
  \caption[]{Zoomed image of intersection between the interface $\Gamma$ and its
    translated version $\Gamma'$. $\phi$ is an angle between tangent lines to
    the interfaces. Surface of dashed area is
    $\frac{\Delta^2}{\sin \phi}$. Vectors $\bm{a}$ and $\bm{b}$ are unit vectors
    parallel to normals to $\Gamma$ and $\Gamma'$ respectively in the point of
    their intersection.}
  \label{fig:fss-zoomed}
\end{figure}
Now define a functional sequence $F_{ss}(\bm{r}; \Delta)$:
\begin{align*}
  F_{ss}(\bm{r}; \Delta) &= \int\int f_\Gamma(\bm{x}; \Delta) f_\Gamma(\bm{x}
  + \bm{r}; \Delta) dx dy \\
  &= \sum_{k=1}^N S_k/\Delta^2
\end{align*}
Here $N$ is a number of intersections of $\Gamma$ with itself translated by a
vector $\bm{r}$ and $S_k$ is a surface of each such intersection (see
\cref{fig:Fss-explained}).
When $\Delta$ tends to zero, the boundary at the point of intersection can
be replaced with a tangent line at the point of intersection
(\cref{fig:fss-zoomed}). We get an expression for a surface of
intersection: $S_k = \frac{\Delta^2}{\sin \phi_k}$ where $\phi_k$ is an
acute angle between tangent lines. Hence, we get the following equation:
\begin{equation}
  F_{ss}(\bm{r}) = \lim_{\Delta \to 0} F_{ss}(\bm{r}; \Delta) =
  \sum_{k=1}^N \frac{1}{\sin \phi_k} \label{eq:fss-2d-sin}
\end{equation}
It's easy to show that
$\lim_{\Delta \to 0} f_\Gamma(\bm{x}; \Delta) = |\nabla \chi_A(\bm{x})|$
and therefore \cref{eq:fss-2d-sin} and \cref{eq:fss} define the same function.

A less illustrative but more practical way to calculate an expression
$\frac{S_k}{\Delta}$ is the following. Take two vectors $\bm{a}$ and $\bm{b}$
parallel to normals to $\Gamma$ and $\Gamma'$ in $k$-th point of their
intersection such as $|a| = |b| = 1$. If $F_{ss}$ is defined at the point of
consideration then $\bm{a} \ne \pm \bm{b}$. The pair $(\bm{a}, \bm{b})$ may
serve as a basis with the following transformation from the basis
$(1,0), (0,1)$:
\begin{align*}
  x' &= f(x, y) = a_x x + a_y y \\
  y' &= g(x, y) = b_x x + b_y y
\end{align*}
where $\bm{a} = (a_x, a_y)$, $\bm{b} = (b_x, b_y)$. Jacobian determinant for
this transformation is
\begin{equation*}
  J = \left|
  \begin{array}{cc}
    \frac{\partial f}{\partial x} & \frac{\partial f}{\partial y} \\
    \frac{\partial g}{\partial x} & \frac{\partial g}{\partial y} \\
  \end{array}
  \right| =
  \left|
  \begin{array}{cc}
    a_x & a_y \\
    b_x & b_y
  \end{array}
  \right| = a_x b_y - a_y b_x
\end{equation*}

@@

Now we get to practical algorithm for computation of $F_{ss}$ for sets with a
smooth boundary. Suppose a set $S$ is defined as
$S = \left\{ (x, y) \ | \quad f(x, y) \le T \right\}$ where $f(x, y)$ is some
function differentiable at points where $f(x, y) = T$ and $T$ is some
threshold. Let's calculate surface-surface function for this set at a point
$\bm{r} = (r_x, r_y)$. We find all intersections between a boundary $\Gamma$ and
its translated version by solving a system of equations
\begin{equation*}
  \left\{
  \begin{array}{l}
    f(x, y) = T \\
    f(x-r_x, y-r_y) = T
  \end{array}
  \right.
\end{equation*}
which can be replaced by one equation
\begin{equation}
  (f(x, y) - T)^2 + (f(x-r_x, y-r_y) - T)^2 = 0 \label{eq:inter}
\end{equation}
We denote a set of solutions of this equation as $X$. For every $\bm{x}_i \in X$
we find normals to the boundary and the boundary translated by a vector
$\bm{r}$. These normals can be found using automatic differentiation and dual
numbers described in \cref{sec:dual}. Then we normalize these normals (i.e. make
their length equal to $1$) and use \cref{eq:fss-2d} to calculate $F_{ss}$.

Summarizing, the algorithm of calculating $F_{ss}(\bm{r})$ is the following:
\begin{enumerate}
\item Find $X$ which is a set of soulutions of \cref{eq:inter}.
\item For all points in $X$ find normals to the boundaries using
  \cref{eq:autonormals} and normalize them.
\item Calculate $F_{ss}(\bm{r})$ using \cref{eq:fss-2d}.
\end{enumerate}

The set $X$ can be found by firstly evaluating $f(x, y)$ in some regular grid
searching for all points $(x_i, y_i)$ for which $|f(x_i, y_i) - T| < \epsilon$
and $|f(x_i - r_x, y_i - r_y) - T| < \epsilon$ for some positive $\epsilon$
and then using these points as starting points for some algorithm based on
gradient descent like ADAM to solve \cref{eq:inter}.

\subsection{Surface-surface-surface function for three-dimensional sets with
  smooth boundary}
\label{sec:fsss-3d}
Results of this section are made by analogy with \cref{sec:fss-2d}. Again, let
$S$ be some set in $\mathbb{R}^3$ which has a smooth boundary, i.e. has a
tangent plane at almost every point of its boundary. Using the function
\cref{eq:delta-sequence} we define a functional sequence
$F_{sss}(\bm{r_1}, \bm{r_2}; \Delta)$:
\begin{align*}
  F_{sss}(\bm{r_1}, \bm{r_2}; \Delta) &= \int\int\int dx dy dz \times \\
  &\qquad \times f_\Gamma(\bm{x}; \Delta) f_\Gamma(\bm{x} + \bm{r_1}; \Delta)
  f_\Gamma(\bm{x} + \bm{r_2}; \Delta) \\
  &= \sum_{k=1}^N V_k/\Delta^3
\end{align*}
$V_k$ is a volume of parallelepiped where all three multiplicands under integral
are non-zero. $N$ is a number of these parallelepipeds, i.e. number of points
where all three boundaries (one ``original'' boundary and two translated
versions) have an intersection. With $\Delta$ going to zero we get a convergence
to $F_{sss}$:
\begin{equation*}
  \lim_{\Delta \to 0} F_{sss}(\bm{r_1}, \bm{r_2}; \Delta) = F_{sss}(\bm{r_1},
  \bm{r_2})
\end{equation*}

Let $\bm{a}$, $\bm{b}$ and $\bm{c}$ be normalized normals to each boundary in
the point of intersection. The corresponding $V_k$ is finite if and only if
$\bm{a} \ne \pm \bm{b} \ne \pm \bm{c}$. In this case we can use a triplet
$(\bm{a}, \bm{b}, \bm{c})$ as basis for $\mathbb{R}^3$ with the following
transformation from the basis $(1,0,0), (0,1,0), (0, 0, 1)$:
\begin{align*}
  x' &= f(x, y) = a_x x + a_y y + a_z z \\
  y' &= g(x, y) = b_x x + b_y y + b_z z \\
  z' &= h(x, y) = c_x x + c_y y + c_z z \\
\end{align*}
Jacobian determinant for this transformation is
\begin{equation*}
  J = \left|
  \begin{array}{ccc}
    a_x & a_y & a_z \\
    b_x & b_y & b_z \\
    c_x & c_y & c_z \\
  \end{array}
  \right|
\end{equation*}
and the expression for $F_{sss}$ becomes
\begin{equation}
  F_{sss}(\bm{r_1}, \bm{r_2}) = \sum_{k=1}^N \frac{1}{|J_k|} \label{eq:fsss-3d}
\end{equation}
where $J_k$ is Jacobian determinant computed at $k$-th point of
intersection.
